---
title: "MovieLens_Report"
author: "Ian Gledhill"
date: "03/06/2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, code = source("Solution.R"), echo=FALSE}
```

\newpage
# Introduction

## Overview
This report documents the process and methodology adopted by this project to devise a movie recommendation system. The movie recommendations are derived from the predicted ratings made by a user for a given movie based on previous ratings made by the user on other films. 
The movie recommendtaion system will analyze a dataset of movie ratings to predict what ratings a particular user would give a movie. 

The users are randomly selected and are only identified by a uniquely assigned user id; there is no demographic and geopgraphic location information used.

## Dataset

### Source
The source dataset was created by the University of Minnesota from the online movie recommendation service MovieLens (https://movielens.org).
 
It is a large dataset and contains:

* 10000054 ratings 

* 95580 tags 

* 10681 movies by 

* 71567 users

The source files can be obtained from this location:

* https://grouplens.org/datasets/movielens/10m/

This link shows further details, including usage rights provided by the license:

* http://files.grouplens.org/datasets/movielens/ml-10m-README.html

### Sample Size for Analysis
This is a very large dataset and for the sake of managability a 10% sample of the source dataset will be used by the project. The source dataset is circa 1M ratings so the sample dataset will be approximately 100K ratings. 

### Shape of the Data

The source data contains three data sets:

* Movies

* Ratings

* Tags

### Training and Test Data Partitions
The full approach is described later in this report but in summary the sample data is split into two partitions.
A *training partition* will be allocated *80%* of the sample data and will be used to create (or _train_) the algorithm. 
A *test partition* will be allocated the remaining *20%* of the sample data and will be used to validate the predictions of the algorithm.


## Goal
The objective of the project was to find a movie recommendation algorithm. Its success will be measured by reference to the RMSE produced when the algorithm is validated against the test dataset. The RMSE will be graded against a predfined set of thresholds to determine the overall quality of the algorithm (described later).



\newpage
# Approach

## Data Cleansing

## Modelling Techniques

The algoritm will start from a naive position and then be refined as further insights into the data are gained. The effect of each refinement will be assessed using a loss function. The RMSE (_residual mean squared error_) will be used as the loss function for this project.

### Version 1
Use the avarage rating of the training data to predict the rating in the test data.

$$
\begin{aligned}
 Y_{u,i} &= \mu + \epsilon_{u,i}
\end{aligned}
$$

### Version 2
Use the mean with the addition of a movie effect to take into account that some movies are rated more favourably then others.

$$
\begin{aligned}
 Y_{u,i} &= \mu + b_i + \epsilon_{u,i}
\end{aligned}
$$

### Version 3
Use the mean with the addition of a user effect to take into account that some users rate movies more generously then others.

$$
\begin{aligned}
 Y_{u,i} &= \mu + b_u + \epsilon_{u,i}
\end{aligned}
$$

### Version 4
Use the mean with the regulated movie and user effect.

$$
\begin{aligned}
 Y_{u,i} &= \mu + b_i + b_u + \epsilon_{u,i}
\end{aligned}
$$


\newpage
# Results
```{r regulated, echo=FALSE}
ggplot(regulated_data, aes(x = lambdas, y = RMSE)) +
    geom_point() +
    geom_line() +
    ggtitle("Regulated Fit", subtitle="Movie and User") +
    xlab("Regulator Value (Lambda)") +
    theme_minimal()
```


```{r methods, echo=FALSE}
ggplot(rmse_results, aes(x = method, y = RMSE)) +
    geom_point() +
    ggtitle("RMSE", subtitle="Changes ") +
    xlab("Method") +
    theme_minimal()
```

\newpage
# Conclusion
```{r pressure2, echo=FALSE}

```

\newpage
# Bibliography and References

